---
title: "Buoy 41002"
author: "Isaac Allen"
date: "October 23, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = 'pdf')
library(tidyr)
library(plyr)
library(knitr)
```

# The Data
## Buoy 41002 South of Cape Hatteras

The data analyzed here is the standard metereological data for the Buoy 225 Nautical Miles South of Cape Hatteras.  This data, especially considering the location, could be useful in weather and hurricane prediction.

The variables consist of two to four letter variable names.  They are documented here:

YY, MM, DD  Year, month and day respectively.
hh, mm      Hour and minute respectively.
WD, WDIR    Wind direction in degrees clockwise from North.
WSPD        Wind speed averaged over an eight-minute period. (m/s)
GST         Peak gust speed in the eight minute period. (m/s)
WVHT        Significant wave height. (m)
DPD         Dominant wave period, the period with maximum wave energy (s)
APD         Average wave period during a 20-minute period (s)
MWD         Wave direction during DPD.  Degrees clockwise from North.
PRES, BAR   Sea level pressure. (hPa)
ATMP        Air temperature. (C)
WTMP        Sea surface temperature. (C)
DEWP        Dewpoint tempurate at the height of ATMP. (C)
VIS         Station visibility. (Nautical Miles)
TIDE        Water level in feet above or below mean lower low water.

More information on the variables can be found on NOAA's website.
http://www.ndbc.noaa.gov/measdes.shtml

# Combining The Files

With each year being its own file, the first step was to save all the files into the "Raw Data" folder, and save them as "YYYY.txt", where YYYY is the year.  Afterwards I created a list of all the file names in an easily reproducible way.

```{r file list, tidy = TRUE}
years <- c(1973:2015)
years <- years[years != 2009 & years != 2011]
files <- paste("Raw Data/", years, ".txt", sep = "")
head(files)
```

# Evaluating Variables

Not all of the data has the same variables in it, so in order to easily evaluate what the variables were and which needed to be changed I created a matrix of the variable names of all the files using the following function. 

From this function I found that BAR = PRES, YYYY = YY, X.YY = YY, #YY = YY, WD = WDIR.  All other variable names are unique.

```{r variable list, tidy = TRUE}
variable_list <- function(file_list){
  variables = matrix(0, nrow = length(file_list), ncol = 18)
  for(i in c(1:length(file_list))){
    names = colnames(read.table(file_list[i], header = TRUE, fill = TRUE, comment.char = ""))
    variables[i,] = c(names, rep("NA", times = 18 - length(names)))
  }
  variables
}
all_variables <- variable_list(files)
tail(all_variables)
```

The function can be made to work for any series of data files, the only thing that would need to be changed is the maximum number of variables, which would involve changing any occurances of 18.

# Combining the Files

The first step in combining the files was to create a helper function that made all of the data files have the same number of variables and the same variables.

```{r same variable, tidy = TRUE}
same_var <- function(df){
  df = rename(df, c("YYYY" = "YY", "X.YY" = "YY", "WD" = "WDIR", "BAR" = "PRES"), warn_missing = FALSE)
  if(length(df) < 18){
    mm = rep(99.0, times = length(df$YY))
    df = cbind(df, mm)
    if(length(df) < 18){
      TIDE = rep(99.00, times = length(df$YY))
      df = cbind(df, TIDE)
    }
  }
  df
}
```

It will add the numeric equivalent of NA values to the added variables and change the previous variable names to a standardized variable name.  I decided to keep "YY", "WDIR", and "PRES" for now.

Next was creating the dataframe that consists of all of the dataframes.

```{r combine, tidy = TRUE,warning= FALSE}
combine_files <- function(file_list){
  full_data <- data.frame()
  for(i in c(1:length(file_list))){
    temp_data = read.table(file_list[i], header = TRUE, fill = TRUE, comment.char = "")
    temp_data = same_var(temp_data)
    full_data = rbind(full_data, temp_data)
  }
  full_data
}

combined_data <- combine_files(files)
(head(combined_data))
```

Each new data file is made into a dataframe and given the same variables with the helper function, and then added to the whole dataframe.

```{r lib, include=FALSE}
library(dplyr)
```

# Removing Large Groups of NA

With columns and rows with only NA values, I needed to remove them and find a way to change the numeric representations of NA values to NA.

First I had to remove the few occurances of text within the data to explain the units used.  Then I had to change the numbers to NA, which were 99, 999, and 9999 depending on the number of digits the maximum value for a column had.  This meant I had to find the maximum value and then choose which number to convert to NA; a column could include 99 or 999 in its range, so I wanted to make sure not to remove any legitimate data.

Finally I removed the columns that had only NA values.  This was all done with the function below, which will work for all dataframes set up completely numerically with the same NA coding.

```{r nona, tidy = TRUE,warning= FALSE}
no_na_var <- function(dataset){
  j = c(1:length(dataset))
  dataset[,j] = lapply(dataset[,j], function(x) as.numeric(as.character(x)))
  dataset = dataset[rowSums(is.na(dataset)) !=ncol(dataset),]
  dataset[is.na(dataset)] = -1
  for(i in c(1:length(dataset))){
    if(max(dataset[,i]) %in% c(99, 999, 9999)){
      dataset[, i][dataset[,i] == max(dataset[,i])] = -1 
    }
  }
  dataset[dataset == -1] <- NA
  dataset = dataset[colSums(!is.na(dataset)) > 0]
  dataset
}

reduced_data <- no_na_var(combined_data)
```

The Tide and the Visability variables were removed, as they only contained NA values.

# Useless Data

I thought there might be some data that gives no information - data with only the date and time marked without any actual information collected.  So I removed rows which only contained date, time, and NA values

```{r nand time, tidy = TRUE}
nand_time <- function(dataset){
  dataset[,1][dataset[,1] < 100] = paste("19", dataset[,1][dataset[,1] < 100], sep = "")
  times = match(c("YY", "MM", "DD", "hh", "mm"), colnames(dataset))
  i = c(1:length(dataset))
  times = subset(i, !(i %in% times))
  new_data = filter(dataset, (rowSums(dataset[, times], na.rm = TRUE)) != 0)
  new_data
}


reduced_rows <- nand_time(reduced_data)
```
Only about 4,000 rows were removed, as shown below by the lengths, but they were unecessary.
```{r, echo = FALSE}
length(reduced_data[,1])
length(reduced_rows[,1])
```

# Combining Time Variables

The hour and minute variables can be combined and so can the year, month, and day. I also fix the date column to look better.

```{r date_time, tidy = TRUE,warning= FALSE}
time_comb <- function(dataset){
  times = match(c("hh", "mm"), colnames(dataset))
  dataset$mm[is.na(dataset$mm)] = "00"
  unite(dataset, "Time", times, sep = ":")
}

date_comb <- function(dataset){
  dates = match(c("YY", "MM", "DD"), colnames(dataset))
  unite(dataset, "Date", dates, sep = "-")
}

reduced_vars <- time_comb(reduced_rows)
reduced_vars <- date_comb(reduced_vars)

time_fix <- function(dataset){
  dataset[,1] = as.Date(dataset[, 1], format = '%Y-%m-%d')
  dataset
}

reduced_vars <- time_fix(reduced_vars)

(head(reduced_vars))
```

# Seperate Files

At this point it seemed like a good time to save and figure out how to best seperate the files.  I created another matrix that contains all of the combinations of variables that exist to look for some kind of pattern.

```{r var combin, tidy = TRUE,warning= FALSE}
var_combinations <- function(dataset){
  i = c(1:length(dataset[,1]))
  mat = dataset 
  mat[!is.na(mat)] = col(mat)[!is.na(mat)]
  mat = 
  mat = distinct(mat)
  mat
}

variable_combos <- var_combinations(reduced_rows)

length(variable_combos[,1])
```

There are 130 rows, meaning 130 different combinations of variables with no distinct patterns for seperation.  Analysis will have to be done by subsetting the information to what is needed, as it would be impractical to create 130 different files for this data.  

## Saving Files

I save both csv and RDS files, and one with the dates and times together and one with them seperate.

```{r save, tidy = TRUE, warning = FALSE}
write.csv(reduced_vars,"Reduced Data - combined time.txt")
write.csv(reduced_rows,"Reduced Data.txt")

saveRDS(reduced_vars,"Reduced Data - combined timeRDS.rds")
saveRDS(reduced_rows,"Reduced DataRDS.rds")
```

# Graphing and Summary Statistics
## Graphs

```{r temp, tidy = TRUE, warning = FALSE, echo = FALSE}
library(ggplot2)
time_tgth <- readRDS("Reduced Data - combined timeRDS.rds")

time_sep <- readRDS("Reduced DataRDS.rds")
temp_by_year <- aggregate(time_sep[,13], list(time_sep$YY), mean, na.rm = TRUE)
ggplot(temp_by_year, aes(Group.1, x)) + geom_point()+ labs(x = "Year", y = "Temp (C)")
```

A graph showing the average temperature versus year.

```{r wind, tidy = TRUE, warning = FALSE, echo= FALSE, fig.height=4, fig.width=4}
wind <- aggregate(time_sep[,6:7], list(time_sep$WDIR), mean, na.rm = TRUE)

ggplot(wind, aes(Group.1, WSPD)) + geom_bar(stat = "identity", col = "blue") + coord_polar(theta = "x", start = 0, direction = 1) + labs(x = "Direction", y = "Wind Speed (m/s)")
```

Average wind strength for each wind direction, pointing in the corresponding direction. North corresponds to 0. 

```{r wind2, tidy = TRUE, warning = FALSE, echo= FALSE, fig.height = 4, fig.width=5}
wind2 <- aggregate(time_sep[,6:7], list(time_sep$MM), mean, na.rm = TRUE)

ggplot(time_sep, aes(as.factor(MM), WSPD)) + geom_boxplot()+ labs(x = "Month", y = "Wind Speed (m/s)")
```

Boxplot of wind speeds based on month.

## Summary Statistics

I created a function to find the mean and variance of all variables not time related.

```{r sum_stat, tidy = TRUE, warning = FALSE}
summary_stat <- function(dataset){
  no_time = subset(colnames(dataset), !(colnames(dataset) %in% c("Date", "Time")))
  mat = matrix(0, nrow = 2, ncol = length(no_time))
  for(i in c(1:length(no_time))){
    mat[1,i] = mean(dataset[,i + 2], na.rm = TRUE)
    mat[2,i] = var(dataset[,i + 2], na.rm = TRUE)
  }
  rownames(mat) <- c("Mean", "Variance")
  colnames(mat) <- no_time
  mat
}

statis <- summary_stat(time_tgth)
(statis)
```